---
title: "Cox Proportional Hazards Model Group Project"
author: "Edwin Alvarado, Darnell Thomas, and Michael Ho"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
toc: true
toc_float: true 
course: STA 6257 - Advance Statistical Modeling
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

## 1. Introduction

The Cox Proportional Hazards (CPH) model is a method used for performing
survival analyses, namely in the medical field. It is used to relate
several risk factors simultaneously against survival time and is widely
considered to be one of the most popular methods in survival analyses.
One of the key reasons why it is popular in survival analysis is due to
its ability to handle censored time-to-event data [@fisher1999time].
"Censored" in this case would mean that some observed units were
recorded for various lengths of time, but the event of interest did not
occur during those studies. With this, Sir David Cox, the inventor of
the model, considered such predictors and covariates when creating the
Cox method. The risks, or "hazards," were thought of as proportional to
the probability of an event occurring at any particular time. Some of
the risk factors used in its analyses include covariates -- independent
variables that can influence the outcome of an event but are not of
direct interest in a study - such as treatment, age, gender, height, and
so on. Another key factor in the CPH's popularity in survival analysis
is its semiparametric nature [@kleinbaum2012cox]. The composition of the
model is as follows:

$$ H(t) = H_0(t) \times exp(b_1x_1 + b_2x_2 + ... + b_kx_k) $$

Where $H_0 (t)$ is considered to be the baseline hazard function of the
model. If in the event all other variables X are equal to zero, the
formula reduces to $H_0 (t)$ ; hence, it is the baseline hazard
function. It is due to the hazard function being unspecified that the
CPH model can be considered semiparametric. In other words, it is
semiparametric because there is no assumption about the distribution of
survival times, but rather assumes that the effects of different
variables that affect survival are constant over time, hence the
composition of its formula [@d2021methods]. Where parametric models have
completely specified functional forms, the unspecified nature of the
baseline hazard function allows the CPH model to be naturally more
flexible. Another reason for its popularity is that even with an
unspecified hazard function, the CPH model is robust, being able to
obtain favorable results such as good estimates of regression
coefficients, hazard ratios of interest, and adjusted survival curves
[@kleinbaum2012cox]. As a robust model, the CPH method can obtain
results comparable to those of a parametric model. While parametric
methods are still preferred over the CPH models in cases where the
correct method is clearly indicated, the CPH model demonstrates its
versatility by being both semiparametric in nature as well as attaining
comparable results to its parametric counterparts.

## 2. Methods

The Cox PH model proposed in 1972 allows the estimation of regression
coefficients of a hazard function without specifying the baseline hazard
[@cox1972regression] The hazard function is the instantaneous risk of
failure at time (t). The hazard function for other survival
distributions such as Weibull take the form below:

$$h_i(t)= exp \times (α+ Z_iβ) $$

where

$h_i(t)$ is the hazard function with respect to time

α is a constant that represents baseline hazard when all Zs are zero

$Z_i$ are the covariates measured for $i$-th individuals
($Z_{1i}+, Z_{2i}, ..., Z_{pi}$) is the unknown regression coefficients
[@fox2002r].

In contrast to Weibull, the Cox PH model leaves the baseline hazard
unspecified (i.e. $α=h_i(t)$ ), and this baseline function can fluctuate
with respect time instead of remaining constant.

$$h_i(t)= h_0(t) exp(Z_iβ) $$ where

$h_i(t)$ is the hazard function with respect to time

$h_0(t)$ is the unknown baseline hazard function which can fluctuate
with respect to time [@cox1972regression].

To demonstrate the impact of differing variables on the hazard function,
Cox used the example of one Z variable (p=1) with values 0 and 1. The
corresponding hazard functions become $h_0(t)$ and
$h_0(t) \times exp(\beta)$ [@cox1972regression]. This example can be
extended to include more variables and used to calculate the hazard
ratio for the two cases:

$$ {h_i(t) \over h_{i'}(t)} = {h_0(t)\times exp(\eta_i) \over h_0(t) \times exp(\eta_{i'})} =  {exp(\eta_i) \over exp(\eta_{i'})} $$

Where

$h_i(t) \over h_{i'}(t)$ is the hazard ratio (HR) and is assumed to be
constant with respect to time [@stensrud2019limitations][@fox2002r]

$\eta_i$ is corresponding linear predictors of $i$

$\eta_{i'}$ is corresponding linear predictors of $i'$

The benefit of the Cox PH model is that the baseline hazard function can
fluctuate with time instead of being constant or multiplicative. Another
benefit is that the regression coefficients can be estimated without
knowing the baseline hazard using the partial likelihood method as
proposed by Cox (1972). Although the baseline hazard function can
fluctuate, the Cox PH model assumes that the hazard ratio (HR) is
constant. If the constant HR assumption is broken, one solution is to
include an interaction between the covariate and time into the Cox model
which is called time-dependent covariates [@fox2002r] . Time dependent
covariates are variables that vary with respect to time. Time dependent
covariates can be modeled in the Cox by replacing $Z_i$ with $Z_i(t)$
which indicates that $Z_i$ varies with respect to time.

$$h_i(t)= h_0(t) exp(Z_i(t)β) $$

where

$h_i(t)$ is the hazard function with respect

$Z_i(t)$ are the covariates that may be time-fixed or time-dependent
[@hendry2014data] is the unknown regression coefficients.

The hazard ratio (HR) then becomes the following:

$$ {h_0(t) \times exp(Z_i(t)β) \over h_0(t) \times exp(Z_j(t)β)} =  {exp(Z_i(t)β) \over exp(Z_j(t)β)} $$

This allows the use of Cox PH for non-constant HR if the covariate is
time-dependent. In general, the log hazard and covariates are assumed to
be linear in the Cox PH model [@fox2002r]. The Cox model also assumes
that the failure time of an individual is independent of the other
individuals. This means that if one participant in a study dies that no
other participants are affected. The Cox PH also assumes that censored
individuals have a failure time that exceeds the censoring time
[@cox1972regression]. For example, a participant is still alive at the
end of the study and is right-censored. The assumptions of proportional
HR, independent individuals and right-censoring also applies to other
exponential survival distributions [@fox2002r] [@hendry2014data]. A key
downside of Cox PH is that the survival function cannot be calculated
since the baseline hazard is unknown. This hinders clinicians in
determining the net benefit of treatments [@mclernon2023assessing].

## 3. Application

The Cox Proportional Hazard model is used for modeling survival data and
identifying relationships between predictors and survival times, and is
commonly used in the medical research field but has also been applied in
fields like finance and the social sciences. [@hendry2014data] In
medical research it is used to identify factors that influence patient
outcomes, e.g. death or other health status. For example,
[@thiruvengadam2021study] used the Cox model to analyze the factors that
determine the length of stay for COVID-19 patients, using different
demographic and health information to find connections to the length of
stay or even death of patients. [@thiruvengadam2021study] Other medical
research often mirrors this structure and the model has also been
applied in the context of clinical trials. For example, researchers used
the model to determine treatment risks and treatment effectiveness for
those in a systolic blood pressure intervention trial, finding a 25%
lower hazard rate for those receiving intensive treatment.
[@stensrud2019limitations] The Cox proportional hazard model excels at
modeling this type of scenario, using survival data that can be
censored, i.e. as in the case where patients have died or left the
study. Although commonly used for medical research, the application is
straightforward in other fields with some research looking at the
duration of events like marriages, recidivism, and bank failures. For
example, researchers examining the factors leading to bank failure used
the 2008 recession as a case study to determine and predict what
factors, e.g. asset ratios and loan amounts, are indicators of failure.
[@cox2017proportional] The model has some limitations especially when
not including time dependent structure in the model. In that case, the
model assumes the hazard ratio is the same across the time sample, which
is often not the case in medical research. This shortcoming is
identified in [@stensrud2020test], which highlights how treatment
effects may not be constant over a whole sample and in some cases may
have no effect for months and then a drastic effect after some time.
This has led to the use of models which can include time dependent
variables.

## 4. Our Dataset

We decided to use a dataset named VA:
<https://vincentarelbundock.github.io/Rdatasets/doc/MASS/VA.html>

The reason for this is because we believe the dataset fulfills the
criteria necessary to conduct a survival analysis - in this case,
studying a veteran's probability of survival from lung cancer. It
contains relevant covariates in a standard Cox model, such as age,
treatment, survival time, and so on.

## 5. Analysis and Results \[\[NEED TO UPDATE FOR OUR MODEL AND ANALYSIS\]\]

### 5.1. Data and Vizualisation

The data used is from a Veteran's Administrative lung cancer trial of
137 male patients with advanced inoperable lung cancer.
@kalbfleisch2011statistical The data set is available through the MASS
package in R.

Variables:

-   stime: survival or follow-up time in days
-   status: dead or alive/censored; 0 (alive), 1 (dead)
-   treat: treatment, either standard or test (chemotherapy); 1
    (standard), 2 (test)
-   age: patient's age in years
-   Karn: Karnofsky score of patient's performance on scale of 0 to 100
-   diag.time: patient's time since diagnosis, measured at time of entry
    to trial, in months
-   cell: one of four cell types; 1 (squamous), 2 (small cell), 3
    (adeno), 4 (large)
-   prior: denotes prior therapy; 10 (yes), 0 (no)

Per the National Cancer Institute, the Karnofsky score is standard way
to measure how well a patient can perform ordinary tasks. A higher score
indicates a higher ability to do these tasks, in our study this score is
taken at the start of the study.

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
library(MASS)
library(GGally)
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
data(VA)
#look at first few rows
kable(head(VA, 3), caption = "Sample Rows from VA Lung Trial Data")

summary_stats <- summary(VA[, c("stime", "status", "age", "Karn", "diag.time")])

#Convert summary statistics to a table
kable(summary_stats, caption = "Summary Statistics for VA Lung Trial Data")

#second prop summary
summary_stats2 <- summary(VA[, c("status", "treat", "cell", "prior")])

#Convert summary statistics to a table
kable(summary_stats2, caption = "Summary Count Statistics for VA Lung Trial Data")
```

```{r data viz}
#Create a scatter plot matrix using the GGally package
va_subset <- VA[, c("stime", "age", "Karn", "diag.time", "treat")]

ggpairs(va_subset, title = "Scatterplot Matrix of VA Lung Trial Data", 
        columns = c("stime", "age", "Karn", "diag.time"), ggplot2::aes(color = treat),
        lower = list(continuous = "points", combo = "dot_no_facet", mapping = ggplot2::aes(color=treat, alpha = 0.8)),
        diag = list(continuous = wrap("densityDiag"), mapping = ggplot2::aes(color = treat, alpha = 0.1))) +
  scale_color_manual(values = c("#00BFC4", "#F8766D"))
```

This plot shows a scatterplot matrix of the variables of Survival Time,
Age, Karnofsky Score, and Diagnosis Time where the lower half is a
typical scatter plot between the two variables in the cell, the diagonal
shows a density plot, and the upper diagonal shows a correlation
coefficient between the two variables with significance indicators. The
plots are grouped/colored by the treatment groups, with those receiving
chemotherapy in red. The only variable significantly correlated with
Survival Time is the Karnofsky score, at the alpha 0.05 level. Looking
at the density plots, the standard group's plot in blue is more
concentrated to the left of the group receiving chemotherapy which
indicates a lower survival time for those in the standard treatment
group. In general survival time seems to be positively correlated with
the other variables, but the strength of this correlation is not
apparent from the graphs.

### 5.2. Statistical Modeling

We will start by using all the variables in the inital Cox PH model.

```{r}
# Import Survival package with Cox proportional-hazard model
library(survival)

# Fit model with all variables.
# stime and status are the output values
cox.mod <- coxph(Surv(stime, status) ~ treat + age + Karn + diag.time + cell + prior, data=VA)

# Concordance Index or Harrel's C (model evaluation: higher is better. 50% is equivalent to guessing.)

# Likelihood ratio useful comparing nested models
summary(cox.mod)
```

Some of the variables had p-values less than the significant level. We
will check if we can drop age, diag.time, and prior from the model
without reducing accuracy.

```{r}
# Can we drop age, diag.time, and prior
cox.mod2 <- coxph(Surv(stime, status) ~ treat + Karn + cell, data=VA)

summary(cox.mod2)
```

We will use ANOVA with the Likelihood Ratio test to determine if we can
simplify the model. Since the difference in Likelihood ratio test is
insignificant, we can simplify the model to treat + Karn + cell.

```{r}
# Using anova with Likelihood ratio test
anova(cox.mod2, cox.mod, test="LRT")

```

```{r}
# Can we also remove treat from the model
cox.mod3 <- coxph(Surv(stime, status) ~ Karn + cell, data=VA)

summary(cox.mod3)

# Using anova with Likelihood ratio test
anova(cox.mod3, cox.mod2, test="LRT")
```

With the other variables removed, we will check if treatment is now a
significant variable in the model using another ANOVA test. Based on the
ANOVA test, we can also remove treatment from the model

$$ h(t) = {h_0(t)} \times exp(-0.031X_{Karn}+0.715X_{cell-small}+1.16X_{cell-adeno}+0.325X_{cell-large}) $$

```{r}
summary(cox.mod3)
```

### 5.3. Interpretations

At a given instant of time, the probability of dying for a male VA
patient as Karnofsky score increases by 1 point is 0.969 times less (or
3.15% lower) after adjusting for cell type.

At a given instant of time, the probability of dying for a male VA
patient with small cell is 2.04 times more than a male VA patient with
squamous cell type after adjusting for Karnofsky score.

At a given instant of time, the probability of dying for a male VA
patient with adeno cell is 3.18 times more than a male VA patient with
squamous cell type after adjusting for Karnofsky score.

At a given instant of time, the probability of dying for a male VA
patient with large cell is 1.38 times more than a male VA patient with
squamous cell type after adjusting for Karnofsky score.

### 5.4. Testing Proportional Assumption

```{r}
# Using cox.zph() to test for proportional-hazards assumption
cox.zph(cox.mod3)
```

```{r}
# Plotting graphs of Schoenfield residuals vs time
par(mfrow=c(1,2))
plot(cox.zph(cox.mod3))
```

Our current model fails the proportional hazard assumption based on
p-values less than 0.05 and can be observed visually.

## 6. Conclusion

Lorem Ipsum

## 7. References
