---
title: "Cox Proportional Hazards Model Group Project"
author: "Edwin Alvarado, Darnell Thomas, and Michael Ho"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: STA 6257 - Advance Statistical Modeling
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
---

## Introduction
The Cox Proportional Hazards (CPH) model is a method used for performing
survival analyses, namely in the medical field. It is used to relate
several risk factors simultaneously against survival time and is widely
considered to be one of the most popular methods in survival analyses.
One of the key reasons why it is popular in survival analysis is due to
its ability to handle censored time-to-event data [@fisher1999time]. "Censored" in this case would mean that some observed
units were recorded for various lengths of time, but the event of
interest did not occur during those studies. With this, Sir David Cox,
the inventor of the model, considered such predictors and covariates
when creating the Cox method. The risks, or "hazards," were thought of
as proportional to the probability of an event occurring at any
particular time. Some of the risk factors used in its analyses include
covariates -- independent variables that can influence the outcome of an
event but are not of direct interest in a study - such as treatment,
age, gender, height, and so on. Another key factor in the CPH's
popularity in survival analysis is its semiparametric nature (Kleinbaum,
D. G., & Klein, M. (2011)). The composition of the model is as follows:

$$ H(t) = H_0(t) \times exp(b_1x_1 + b_2x_2 + ... + b_kx_k) $$

Where $H_0 (t)$ is considered to be the baseline hazard function of the
model. If in the event all other variables X are equal to zero, the
formula reduces to $H_0 (t)$ ; hence, it is the baseline hazard function.
It is due to the hazard function being unspecified that the CPH model
can be considered semiparametric. In other words, it is semiparametric
because there is no assumption about the distribution of survival times,
but rather assumes that the effects of different variables that affect
survival are constant over time, hence the composition of its formula
[@d2021methods]. Where parametric
models have completely specified functional forms, the unspecified
nature of the baseline hazard function allows the CPH model to be
naturally more flexible. Another reason for its popularity is that even
with an unspecified hazard function, the CPH model is robust, being able
to obtain favorable results such as good estimates of regression
coefficients, hazard ratios of interest, and adjusted survival curves
(Kleinbaum, D. G., & Klein, M. (2011)). As a robust model, the CPH
method can obtain results comparable to those of a parametric model.
While parametric methods are still preferred over the CPH models in
cases where the correct method is clearly indicated, the CPH model
demonstrates its versatility by being both semiparametric in nature as
well as attaining comparable results to its parametric counterparts.


## Methods


The Cox PH model proposed in 1972 allows the estimation of regression
coefficients of a hazard function without specifying the baseline hazard
[@cox1972regression] The hazard function is the instantaneous risk of failure at time
(t). The hazard function for other survival distributions such as
Weibull take the form below: 

$$h_i(t)= exp \times (α+ Z_iβ) $$

where 

$h_i(t)$ is the hazard function with respect to time 

α is a constant that represents baseline hazard when all Zs are zero 

$Z_i$ are the covariates measured for $i$-th individuals ($Z_{1i}+, Z_{2i}, ..., Z_{pi}$) is the unknown regression coefficients [@fox2002r].

In contrast to Weibull, the Cox PH model leaves the baseline hazard
unspecified (i.e. $α=h_i(t)$ ), and this baseline function can fluctuate with
respect time instead of remaining constant. 

$$h_i(t)= h_0(t) exp(Z_iβ) $$
where 

$h_i(t)$ is the hazard function with respect to time 

$h_0(t)$ is the unknown baseline hazard function which can fluctuate with respect to time [@cox1972regression]. 

To demonstrate the impact of
differing variables on the hazard function, Cox used the example
of one Z variable (p=1) with values 0 and 1. The corresponding hazard
functions become $h_0(t)$ and $h_0(t) \times exp(\beta)$ [@cox1972regression]. This example can be extended to
include more variables and used to calculate the hazard ratio for the
two cases:

$$ {h_i(t) \over h_{i'}(t)} = {h_0(t)\times exp(\eta_i) \over h_0(t) \times exp(\eta_{i'})} =  {exp(\eta_i) \over exp(\eta_{i'})} $$

Where 

$h_i(t) \over h_{i'}(t)$ is the hazard ratio (HR) and is assumed to be constant with
respect to time [@stensrud2019limitations][@fox2002r] 

$\eta_i$ is corresponding linear predictors of $i$

$\eta_{i'}$ is corresponding linear predictors of $i'$ 

The benefit of the Cox PH model
is that the baseline hazard function can fluctuate with time instead of
being constant or multiplicative. Another benefit is that the regression
coefficients can be estimated without knowing the baseline hazard using
the partial likelihood method as proposed by Cox (1972). Although the
baseline hazard function can fluctuate, the Cox PH model assumes that
the hazard ratio (HR) is constant. If the constant HR assumption is
broken, one solution is to include an interaction between the covariate
and time into the Cox model which is called time-dependent covariates
[@fox2002r] . Time dependent covariates are variables that vary with respect
to time. Time dependent covariates can be modeled in the Cox by
replacing $Z_i$ with $Z_i(t)$ which indicates that $Z_i$ varies with respect to
time. 

$$h_i(t)= h_0(t) exp(Z_i(t)β) $$

where 

$h_i(t)$ is the hazard function with respect

$Z_i(t)$ are the covariates that may be time-fixed or
time-dependent [@hendry2014data] is the unknown regression coefficients. 

The hazard ratio (HR) then becomes the following:

$$ {h_0(t) \times exp(Z_i(t)β) \over h_0(t) \times exp(Z_j(t)β)} =  {exp(Z_i(t)β) \over exp(Z_j(t)β)} $$

This allows the use of Cox PH for non-constant HR if the covariate is time-dependent. In
general, the log hazard and covariates are assumed to be linear in the
Cox PH model [@fox2002r]. The Cox model also assumes that the failure time of
an individual is independent of the other individuals. This means that
if one participant in a study dies that no other participants are
affected. The Cox PH also assumes that censored individuals have a
failure time that exceeds the censoring time [@cox1972regression]. For example, a
participant is still alive at the end of the study and is
right-censored. The assumptions of proportional HR, independent
individuals and right-censoring also applies to other exponential
survival distributions [@fox2002r] [@hendry2014data]. A key downside of Cox PH is that
the survival function cannot be calculated since the baseline hazard is
unknown. This hinders clinicians in determining the net benefit of
treatments [@mclernon2023assessing].

## Application

The Cox Proportional Hazard model is used for modeling survival data and
identifying relationships between predictors and survival times, and is
commonly used in the medical research field but has also been applied in
fields like finance and the social sciences. [@hendry2014data] In medical
research it is used to identify factors that influence patient outcomes,
e.g. death or other health status. For example, [@thiruvengadam2021study] used the Cox model to analyze the factors that determine the
length of stay for COVID-19 patients, using different demographic and
health information to find connections to the length of stay or even
death of patients. [@thiruvengadam2021study] Other medical research often mirrors
this structure and the model has also been applied in the context of
clinical trials. For example, researchers used the model to determine
treatment risks and treatment effectiveness for those in a systolic
blood pressure intervention trial, finding a 25% lower hazard rate for
those receiving intensive treatment. [@stensrud2019limitations] The Cox proportional
hazard model excels at modeling this type of scenario, using survival
data that can be censored, i.e. as in the case where patients have died
or left the study. Although commonly used for medical research, the
application is straightforward in other fields with some research
looking at the duration of events like marriages, recidivism, and bank
failures. For example, researchers examining the factors leading to bank
failure used the 2008 recession as a case study to determine and predict
what factors, e.g. asset ratios and loan amounts, are indicators of
failure. [@cox2017proportional] The model has some limitations especially
when not including time dependent structure in the model. In that case,
the model assumes the hazard ratio is the same across the time sample,
which is often not the case in medical research. This shortcoming is
identified in [@stensrud2020test], which highlights how treatment effects
may not be constant over a whole sample and in some cases may have no
effect for months and then a drastic effect after some time. This has
led to the use of models which can include time dependent variables.


## Our Dataset

We decided to use a dataset named VA:
<https://vincentarelbundock.github.io/Rdatasets/doc/MASS/VA.html>

The reason for this is because we believe the dataset fulfills the
criteria necessary to conduct a survival analysis - in this case,
studying a veteran's probability of survival from lung cancer. It
contains relevant covariates in a standard Cox model, such as age,
treatment, survival time, and so on.

## Analysis and Results [[NEED TO UPDATE FOR OUR MODEL AND ANALYSIS]]
### Data and Vizualisation
A study was conducted to determine how...


```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```


```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_wsj()
  

```

### Statistical Modeling

### Conclusion

## References
