---
title: "Cox Proportional Hazards Model Group Project"
author: "Edwin Alvarado, Darnell Thomas, and Michael Ho"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
toc: true
toc_float: true 
fig_caption: yes 
course: STA 6257 - Advance Statistical Modeling
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

[Slides](https://e-alvarado.github.io/STA6257_coxph/slides#/title-slide)

## 1. Introduction

The Cox Proportional Hazards (CPH) model is a method used for performing
survival analyses, often in the medical field. It is used to relate
several risk factors simultaneously against survival time and is widely
considered to be one of the most popular methods in survival analyses.
One of the key reasons why it is popular in survival analysis is due to
its ability to handle censored time-to-event data [@fisher1999time].
"Censored" in this case would mean that some observed units were
recorded for various lengths of time, but the event of interest did not
occur during those studies. In our specific case-study, the event of interest would mean the patient passing away. With this, Sir David Cox, the inventor of
the model, considered such predictors and covariates when creating the
Cox method. The risks, or "hazards," were thought of as proportional to
the probability of an event occurring at any particular time. Some of
the risk factors used in its analyses include covariates -- independent
variables that can influence the outcome of an event but are not of
direct interest in a study - such as treatment, age, gender, height, and
so on. Another key factor in the CPH's popularity in survival analysis
is its semiparametric nature [@kleinbaum2012cox]. The composition of the
model is as follows:

$$ H(t) = H_0(t) \times exp(b_1x_1 + b_2x_2 + ... + b_kx_k) $$

Where $H_0 (t)$ is considered to be the baseline hazard function of the
model. If in the event all other variables X are equal to zero, the
formula reduces to $H_0 (t)$ ; hence, it is the baseline hazard
function. It is due to the hazard function being unspecified that the
CPH model can be considered semiparametric. In other words, it is
semiparametric because there is no assumption about the distribution of
survival times, but rather assumes that the effects of different
variables that affect survival are constant over time, hence the
composition of its formula [@d2021methods]. Where parametric models have
completely specified functional forms, the unspecified nature of the
baseline hazard function allows the CPH model to be naturally more
flexible. Another reason for its popularity is that because of the
hazard function being unspecified, the CPH model is robust, being able to
obtain favorable results such as good estimates of regression
coefficients, hazard ratios of interest, and adjusted survival curves
[@kleinbaum2012cox]. As a robust model, the CPH method can obtain
results comparable to those of a parametric model. While parametric
methods are still preferred over the CPH models in cases where the
correct method is clearly indicated, the CPH model demonstrates its
versatility by being both semiparametric in nature as well as attaining
comparable results to its parametric counterparts.

### 1.1. Application

While the Cox Proportional Hazard model is mainly used for modeling
survival data and identifying relationships between predictors and
survival times, and is commonly used in the medical research field, it
has also been applied in fields like finance and the social sciences.
[@hendry2014data] In medical research it is used to identify factors
that influence patient outcomes, e.g. death or other health status. For
example, [@thiruvengadam2021study] used the Cox model to analyze the
factors that determine the length of stay for COVID-19 patients, using
different demographic and health information to find connections to the
length of stay or even death of patients. [@thiruvengadam2021study]
Other medical research often mirrors this structure and the model has
also been applied in the context of clinical trials. For example,
researchers used the model to determine treatment risks and treatment
effectiveness for those in a systolic blood pressure intervention trial,
finding a 25% lower hazard rate for those receiving intensive treatment.
[@stensrud2019limitations] The Cox proportional hazard model excels at
modeling this type of scenario, using survival data that can be
censored, i.e. as in the case where patients have died or left the
study.

Although commonly used for medical research, the application is
straightforward in other fields with some research looking at the
duration of events like marriages, recidivism, and bank failures. For
example, researchers examining the factors leading to bank failure used
the 2008 recession as a case study to determine and predict what
factors, e.g. asset ratios and loan amounts, are indicators of failure.
[@cox2017proportional]

The model has some limitations especially when not including time
dependent structure in the model. In that case, the model assumes the
hazard ratio is the same across the time sample, which is often not the
case in medical research. This shortcoming is identified in
[@stensrud2020test], which highlights how treatment effects may not be
constant over a whole sample and in some cases may have no effect for
months and then a drastic effect after some time. This has led to the
use of models which can include time dependent variables.

## 2. Methods

The Cox PH model proposed in 1972 allows the estimation of regression
coefficients of a hazard function without specifying the baseline hazard
[@cox1972regression] The hazard function is the instantaneous risk of
failure at time (t). The hazard function for other survival
distributions such as Weibull take the form below:

$$h_i(t)= exp (α+ Z_iβ) $$

where

$h_i(t)$ is the hazard function with respect to time

α is a constant that represents baseline hazard when all Zs are zero

$Z_i$ are the covariates measured for $i$-th individuals
($Z_{1i}+, Z_{2i}, ..., Z_{pi}$) is the unknown regression coefficients
[@fox2002r].

In contrast to Weibull, the Cox PH model leaves the baseline hazard
unspecified (i.e. $α=h_i(t)$ ), and this baseline function can fluctuate
with respect time instead of remaining constant.

$$h_i(t)= h_0(t) exp(Z_iβ) $$ where

$h_i(t)$ is the hazard function with respect to time

$h_0(t)$ is the unknown baseline hazard function which can fluctuate
with respect to time [@cox1972regression].

To demonstrate the impact of differing variables on the hazard function,
Cox used the example of one Z variable (p=1) with values 0 and 1. The
corresponding hazard functions become $h_0(t)$ and
$h_0(t) \times exp(\beta)$ [@cox1972regression]. This example can be
extended to include more variables and used to calculate the hazard
ratio for the two cases:

$$ HR = {h_i(t) \over h_{i'}(t)} = {h_0(t)\times exp(\eta_i) \over h_0(t) \times exp(\eta_{i'})} =  {exp(\eta_i) \over exp(\eta_{i'})} $$

Where

$h_i(t) \over h_{i'}(t)$ is the hazard ratio (HR) and is assumed to be
constant with respect to time [@stensrud2019limitations][@fox2002r]

$\eta_i$ is corresponding linear predictors of $i$

$\eta_{i'}$ is corresponding linear predictors of $i'$

The benefit of the Cox PH model is that the baseline hazard function can
fluctuate with time instead of being constant or multiplicative. Another
benefit is that the regression coefficients can be estimated without
knowing the baseline hazard using the partial likelihood method as
proposed by Cox [@cox1972regression]. Although the baseline hazard function can
fluctuate, the Cox PH model assumes that the hazard ratio (HR) is
constant. If the constant HR assumption is broken, one solution is to
include an interaction between the covariate and time into the Cox model
which is called time-dependent covariates [@fox2002r] . Time dependent
covariates are variables that vary with respect to time. Time dependent
covariates can be modeled in the Cox by replacing $Z_i$ with $Z_i(t)$
which indicates that $Z_i$ varies with respect to time.

$$h_i(t)= h_0(t) \times exp(Z_i(t)β) $$

where

$h_i(t)$ is the hazard function with respect

$Z_i(t)$ are the covariates that may be time-fixed or time-dependent
[@hendry2014data] 

$\beta$ are the unknown regression coefficients.

The hazard ratio (HR) then becomes the following:

$$ {h_0(t) \times exp(Z_i(t)β) \over h_0(t) \times exp(Z_j(t)β)} =  {exp(Z_i(t)β) \over exp(Z_j(t)β)} $$

This allows the use of Cox PH for non-constant HR if the covariate is
time-dependent. A similar approach can be used for time-independent covariates that violate proportional hazard assumption. In general, the log hazard and covariates are assumed to
be linear in the Cox PH model [@fox2002r]. The Cox model also assumes
that the failure time of an individual is independent of the other
individuals. This means that if one participant in a study dies that no
other participants are affected. The Cox PH also assumes that censored
individuals have a failure time that exceeds the censoring time
[@cox1972regression]. For example, a participant is still alive at the
end of the study and is right-censored. The assumptions of proportional
HR, independent individuals and right-censoring also applies to other
exponential survival distributions [@fox2002r] [@hendry2014data]. A key
downside of Cox PH is that the survival function cannot be calculated directly and is estimated
since the baseline hazard is unknown. This hinders clinicians in
determining the net benefit of treatments [@mclernon2023assessing].

## 3. Analysis and Results

### 3.1. Purpose

The focus of this research is to find out if the treatment has a
positive effect on survival time, and we will test if any of the other
variables also have any significant effect on survival time. The
Proportional Hazards (PH) assumption must also remain valid to prevent
biased effect estimates.

### 3.2. Data Description

The data used is from a Veteran's Administrative lung cancer trial of
137 male patients with advanced inoperable lung cancer.
[@kalbfleisch2011statistical] The data set is available through the MASS
package [@mass] in R [@rproject]. This data is ideal for modeling with a Cox Proportional
Hazard model with the survival time and treatment variables being of the
most interest.

```{r data description, echo=FALSE, warning=FALSE, results = "asis"}
library(kableExtra)
a <- data.frame(vars = c('stime', 'status', 'treat', 'age', 'Karn', 'diag', 'cell', 'prior'),
               
                DataType = c('Numeric', 'Numeric', 'Nominal', 'Numeric', 'Numeric', 'Numeric','Nominal','Nominal'),
                Description = c('survival or follow-up time, in days',
    'dead or alive/censored; 0 (alive), 1 (dead)',
    'treatment, either standard or test (chemotherapy); 1 (standard), 2 (test)',
    'patient age in years',
    'Karnofsky score of patient performance, on scale of 0 to 100',
    'patient time since diagnosis, measured at time of entry to trial, in months',
    'one of four cell types; 1 (squamous), 2 (small cell), 3 (adeno), 4 (large)',
    'denotes prior therapy; 10 (yes), 0 (no)'),
                stringsAsFactors = FALSE)
 
kable(a,
      caption = "Table 1: Data Definition",
      col.names = c("Attribute", "Type", "Description"),
      align="lcl", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = T, background = "#666666", color = "white")
```

Per the National Cancer Institute, the Karnofsky score is standard way
to measure how well a patient can perform ordinary tasks. A higher score
indicates a higher ability to do these tasks, in our study this score is
taken at the start of the study.

### 3.3. Data Visualization

```{r load libraries, warning=FALSE, echo=F, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
library(MASS)
library(GGally)
library(gtsummary)

```

```{r, warning=FALSE, echo=TRUE}
# Load Data
data(VA)
#look at first few rows
kable(head(VA, 3), caption = "Table 2: Sample Rows from VA Lung Trial Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = T, background = "#666666", color = "white")

summary_stats <- summary(VA[, c("stime", "age", "Karn", "diag.time")])

#Convert summary statistics to a table
kable(summary_stats, caption = "Table 3: Summary Statistics for VA Lung Trial Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = T, background = "#666666", color = "white")


# Copied VA dataset so we can safely convert status to factors
VA.summary <- VA
VA.summary$status <- as.factor(VA.summary$status)

#second prop summary
summary_stats2 <- summary(VA.summary[, c("status", "treat", "cell", "prior")])

#Convert summary statistics to a table
kable(summary_stats2, caption = "Table 4: Summary Count Statistics for VA Lung Trial Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = T, background = "#666666", color = "white")
```

```{r data viz}
#Create a scatter plot matrix using the GGally package
va_subset <- VA[, c("stime", "age", "Karn", "diag.time", "treat")]

ggpairs(va_subset, title = "Figure 1: Scatterplot Matrix of VA Lung Trial Data",
        columns = c("stime", "age", "Karn", "diag.time"), ggplot2::aes(color = treat),
        lower = list(continuous = "points", combo = "dot_no_facet", mapping = ggplot2::aes(color=treat, alpha = 0.8)),
        diag = list(continuous = wrap("densityDiag"), mapping = ggplot2::aes(color = treat, alpha = 0.1))) +
  scale_color_manual(values = c("#00BFC4", "#F8766D")) +
  labs(subtitle = "Grouped by Treatment, Standard (Blue) and Test (Red)")
```

This plot shows a scatterplot matrix of the variables of Survival Time,
Age, Karnofsky Score, and Diagnosis Time where the lower half is a
typical scatter plot between the two variables in the cell, the diagonal
shows a density plot, and the upper diagonal shows a correlation
coefficient between the two variables with significance indicators. One asterisk denotes significance at an alpha = 0.10 level, ** = 0.05 level, and *** = <0.001 level. The plots are grouped/colored by the treatment groups, with those receiving
the test treatment in red, and those receiving the standard treatment in blue. The only variable significantly correlated with
Survival Time is the Karnofsky score, at the alpha 0.05 level. Looking
at the density plots, the standard group's plot in blue is more
concentrated to the left of the group receiving test treatment which
indicates a lower survival time for those in the standard treatment
group. In general, age and the diagnosis time variables, do not seem to be clearly correlated with survival time.  The Karnofsky score seems to be positively correlated with the survival time, which is reinforced by the significant correlation between the two. 
We might expect Karnofsky score to then be correlated in the models, but this might change when accounting for the other variables. 

### 3.4. Statistical Modeling

#### Treatment Only Cox PH Model

Cox PH model allows us to determine the effect of variables by using the
hazard ratio (HR). First, we will use treatment variable to determine if
it is significant. We will use the survival package [@survival-package] in R to run the Cox PH.

```{r, echo = FALSE}
# Import Survival package with Cox proportional-hazard model
library(survival)
```

```{r}
# Label variables for gtsummary tables
attr(VA$treat, "label") = "Treatment"
attr(VA$Karn, "label") = "Karnofsky score"
attr(VA$age, "label") = "Patient's age"
attr(VA$diag.time , "label") = "Time since diagnosis"
attr(VA$cell , "label") = "Cell type"
attr(VA$prior, "label") = "Prior therapy"

levels(VA$treat) = c("standard", "test")
levels(VA$cell) = c("1-squamous", "2-small cell", "3-adeno", "4-large")
levels(VA$prior) = c("yes", "no")
```


```{r}
# Fit model with treatment only
Y = Surv(VA$stime, VA$status)
cox.unimod <- coxph(Y ~ treat, data=VA)
cox.unimod %>% tbl_regression(exponentiate = T) %>% 
add_glance_table(
  include = c(logLik, concordance)
) %>%
  modify_caption("**Table 5: Treatment Only Model**") 

```

The Cox PH model found that the treatment type only is insignificant. A
key assumption of the Cox PH model is that the hazard is proportional
(i.e. the HR is constant across time). To validate the proportional
hazard assumption, we will use the scaled Schoenfeld's residuals to
compare the difference between the observed treatment value and the
weighted-average of other individuals. [@grambsch1994proportional] If
the HR is constant, the smoothed residual curve will stay around 0 as
time increases, which validates the proportional hazard assumption.
[@therneau2017using]

```{r}
# Plotting graphs of Schoenfeld residuals vs time
phtest.unimod = cox.zph(cox.unimod, transform = rank)
plot(phtest.unimod, main = "Figure 2: Schoenfeld Residuals Over Time")
abline(h=0)
```

From the plot, we can see that the confidence interval for the residuals
stay around 0. We will also run a statistical test against the
Schoenfeld residuals to determine if the residuals are correlated with
time. A small p-value (less than 0.05) means there is a significant
correlation between the residuals and time, which means the proportional
hazard assumption is invalid.

```{r}
# Using Scheonfeld to test for proportional-hazards assumption
kable(phtest.unimod$table, caption = "Table 6: Scheonfeld Test for Proportional-Hazards Assumption") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = T, background = "#666666", color = "white")
```

Based on the p-value for treatment, we fail to reject the null
hypothesis that the proportional hazard assumption is valid. Since the
p-value is very close to the cutoff, we will use a graphical method
based the survival probability curve.

We could use Cox PH to estimate survival probability curve based on
fitted model. However, we will use the Kaplan-Meier method, which was
does not depend on a fitted model, to estimate the probability of
survival with respect to time. Using the Kaplan-Meier estimates, we can
plot estimated survivor curves for both treatments. [@therneau2017using] [@kleinbaum2012cox]

```{r}
# Plotting graph of Kaplan-Meier curve
kmfit = survfit(Y ~ treat, data=VA)

plot(kmfit, main = "Figure 3: Survival Curve", lty = c("solid", "dashed"), col=c("black","grey"),
xlab="survival time",ylab="survival probabilities")
legend("topright", c("Standard","Test"), lty=c("solid","dashed"),
col=c("black","grey"))

```

```{r}
#Log log of survival curve
plot(kmfit,fun="cloglog", lty = c("solid", "dashed"), col=c("black","grey") ,xlab="time",ylab="log-log survival", 
     main="Figure 4: Log-Log Curves")
legend("topleft", c("Standard","Test"), lty=c("solid","dashed"),
col=c("black","grey"))
```

```{r}
tbl_survfit(kmfit, 
            times=c(15, 30, 90,180,360),
            label_header="**{time} Days**"
            ) %>%
  modify_caption("**Table 7: Survival Probability for Particular Days for Treatment Only Model**") 
```

From the two plot, we observe that the effectiveness of test treatment
is less effective in the beginning. Then, the test treatment becomes
more effective than the standard treatment towards the end. With
Kaplan-Meier survival curves, we are able to visualize survival
probabilities for both treatment groups. Since the hazard function is
the rate of change in the survival curve, we can use the Kaplan-Meier to
also validate the proportional hazard assumption. The survival curves
would be parallel if the hazard was proportional. However, the curves
cross multiple time which informs us that the proportional hazard
assumption is actually invalid.

#### Multiple Variable Cox PH Model

Now, we will insert all the variables in the Cox PH model and use
Schoenfeld residuals to determine which variables specifically violate
the proportional hazard assumption.

```{r}
# Fit model with all variables.
# stime and status are the output values
cox.mod <- coxph(Y ~ treat + age + Karn + diag.time + cell + prior, data=VA)

cox.mod %>% tbl_regression(exponentiate = T) %>% 
add_glance_table(
  include = c(logLik, concordance)
) %>%
  modify_caption("**Table 8: Multiple Variable Cox PH Model**") 
```

```{r}
# Using cox.zph() to test for proportional-hazards assumption
phtest.mod = cox.zph(cox.mod, transform = rank)

kable(phtest.mod$table, caption = "Table 9: Proportional-Hazards Assumption Test") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = T, background = "#666666", color = "white")
```

```{r}
# Plotting graphs of Schoenfeld residuals vs time
par(mfrow=c(1,2))
plot(phtest.mod[3])
plot(phtest.mod[5])
mtext("Figure 5: Schoenfeld Residuals Over Time", side = 3, line = -3, outer = TRUE)

```

The significant variables of Karnofsky score and cell type were found to
violate the proportional hazard assumption. In order to use them for
analysis, we could only use Cox PH for a portion of the data, implement
a Stratified Cox model, or use an Extended Cox PH model for Time-Varying
Coefficients.

#### Cox PH Model on events after 180 days

We try to filter the data to events greater than 180 days and run the
Cox PH model for that time interval only. The idea is that the hazard
ratio will be constant over that time period.

```{r}
#Filter to only include events after 180 days
VA.180 = filter(VA, stime >= 180)
tbl_summary(VA.180) %>%
  modify_caption("**Table 10: Summary Statistics For Stime Over 180 Days**")

summary_stats3 <- summary(VA.180[, c("stime", "status", "age", "Karn", "diag.time")])
```

After limiting the dataset to only greater 180 days, we are only left
with 27 events. This sample size is too small to make any definitive
conclusions as you will see in the test statistics below.

```{r}
# Fit model with all variables.
# stime and status are the output values
cox.mod180 <- coxph(Surv(stime, status) ~ treat + age + Karn + diag.time + cell + prior, data=VA.180)

cox.mod180 %>% tbl_regression(exponentiate = T) %>% 
add_glance_table(
  include = c(logLik, concordance)
) %>%
  modify_caption("**Table 11: Events after 180 days Model**") 
```

We will use ANOVA with the Likelihood Ratio test to determine if we can
simplify the model.

```{r}
cox.mod180.r <- coxph(Surv(stime, status) ~ treat + diag.time, data=VA.180)

kable(anova(cox.mod180.r, cox.mod180, test="LRT"), caption = "Table 12: ANOVA Test On >180 Days Models") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = T, background = "#666666", color = "white")
cox.mod1 = cox.mod180
```

Although we may be tempted to remove the insignificant variables, the
model performs worse with the reduced model having a significantly lower
Log Likelihood. [@kleinbaum2012cox] Below is the final model using this method.

$$h(t) = {h_0(t)} \times \exp(-0.317 \text{(treat-test)}+0.012 \text{(age)}-0.0087 \text{(Karn)}+0.112\text{(diag.time)}-0.502 \text{(cell-small)}+0.965 \text{(cell-large)}-0.642 \text{(prior-no)})$$

```{r}
# Using cox.zph() to test for proportional-hazards assumption
phtest.mod = cox.zph(cox.mod180, transform = rank)

kable(phtest.mod$table, caption = "Table 13: Proportional-Hazards Assumption Test On >180 Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = T, background = "#666666", color = "white")
```

Since this model is not likely to produce significant results, we will
look at using the Stratified Cox model on the full dataset.

#### Stratified Cox model

The Stratified Cox model assumes a different baseline hazard function
for each categorical variable that does not meet the proportional hazard
assumption. We can stratify on the non-proportional cell type, which
will estimate the effects of the other variables for each cell type. The
downside is that we cannot compare the effect of each cell type on
survival using hazard ratios (HR).

```{r}
cox.modstrat <- coxph(Y ~ treat + age + Karn + diag.time + strata(cell) + prior, data=VA)
cox.modstrat %>% tbl_regression(exponentiate = T) %>% 
add_glance_table(
  include = c(logLik, concordance)
) %>%
  modify_caption("**Table 14: Stratified Cox Model**") 
```

This form of the stratified Cox model assumes that the coefficients are
the same across each of the four (4) cell types. We can also run a
stratified Cox model where the coefficients can vary between strata by
using interaction terms. [@kleinbaum2012cox]

```{r}
cox.modstratI <- coxph(Y ~ treat + age + Karn + diag.time + prior + treat:cell + age:cell + Karn:cell + diag.time:cell +strata(cell) + prior:cell, data=VA)
#anova(cox.modstrat, cox.modstratI, test="LRT")
kable(anova(cox.modstrat, cox.modstratI, test="LRT"), caption = "Table 15: ANOVA Test On Stratified Models") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = T, background = "#666666", color = "white")
```

Based on the Likelihood Ratio test, we can assume that the
no-interaction stratified Cox model is sufficient. Below is the final
model:

$$h_g(t) = {h_{0g}(t)} \times \exp(0.286\text{(treat-test)}-0.012\text{(age)}-0.0383\text{(Karn)}-0.0034\text{(diag.time)}+0.169\text{(prior-no)})$$
where $$g = 1, 2, 3, 4 \text{ (number of cell types)}$$

```{r}
cox.mod2 = cox.modstrat
phtest.mod2 = cox.zph(cox.modstrat, transform=rank)
kable(phtest.mod2$table, caption = "Table 16: Proportional-Hazards Assumption Test On Stratified Cox Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = T, background = "#666666", color = "white")
```

The Karnofsky score still violates the proportional hazards
assumption. If we wanted to stratify it, we would first need to combine
the scores into fewer groups. If we grouped Karnofsky
scores into high, medium, and low bins, that would result in 12
different curves. We will not apply stratification to Karnofsky score
in this paper. Below are survival curves for cell type based the
Stratified Cox model.

```{r}
plot(survfit(cox.modstrat), main = "Figure 7: Survival Probabilities by Cell Type Across Survival Time",
     lty = c("solid","dashed", "dotted","longdash"), col=c("black","grey","blue","orange"),
xlab="survival time",ylab="survival probabilities")
legend("topright", c("1-squamous","2-small cell", "3-adeno", "4-large"), lty=c("solid","dashed", "dotted","longdash"),
col=c("black","grey","blue","orange"))
```

#### Extended Cox PH model for Time-Varying Coefficients

Another approach to variables with nonproportional hazard ratios is
specify a function for the coefficients that change with time. This
function could be linear, logarithmic, or a step function. For this
analysis, we will run the Cox PH model over specific time intervals
using a step function for the Karnofsky score. Before we can run the
Cox PH across the intervals of 0-90, 90-180, and 180+ days, we must
create a new column from the original dataset.

```{r}
VA.cp = survSplit(Surv(stime, status)  ~ ., data=VA, cut=c(90,180), episode="tgroup")

#look at first few rows
kable(head(VA.cp, 5), caption = "Table 17: Sample Rows With New Column") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = T, background = "#666666", color = "white")
```

Note the addition of the "tstart" and "tgroup" columns in the new
dataset

```{r}
cox.modextend = coxph(Surv(tstart, stime, status) ~ treat + age + diag.time + cell + prior + Karn:strata(tgroup),
                 data=VA.cp)

cox.modextend %>% tbl_regression(exponentiate = T) %>% 
add_glance_table(
  include = c(logLik, concordance)
) %>%
  modify_caption("**Table 18: Extended Cox Model**") 
```

```{r}
phtest.modex = cox.zph(cox.modextend, transform=rank)
kable(phtest.modex$table, caption = "Table 19: Proportional-Hazards Assumption Test On Extended Cox Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = T, background = "#666666", color = "white")
```

To address the nonproportional hazard of the cell type, we will also
stratify on the cell type to create an stratified-extended Cox model.
[@ratnaningsih2019stratified]

```{r}
cox.modse = coxph(Surv(tstart, stime, status) ~ treat + age + diag.time + strata(cell) + prior + Karn:strata(tgroup),
                 data=VA.cp)

phtest.modse = cox.zph(cox.modse, transform=rank)
kable(phtest.modse$table, caption = "Table 20: Proportional-Hazards Assumption Test On Stratified-Extended Cox Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = T, background = "#666666", color = "white")
```

This model satisfies the proportional hazard assumption. We will now
check if we can remove some of the insignificant variables.

```{r}
cox.modse_r = coxph(Surv(tstart, stime, status) ~ treat + strata(cell) + Karn:strata(tgroup),
                 data=VA.cp)
# Using anova with Likelihood ratio test
kable(anova(cox.modse_r, cox.modse, test="LRT"), caption = "Table 21: ANOVA Test On Stratified-Extended Cox Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = T, background = "#666666", color = "white")
cox.mod3 = cox.modse_r

```

Since the difference in Likelihood ratio test is insignificant, we can
simplify the model to treat + Karn + cell.

```{r}
cox.modse_r %>% tbl_regression(exponentiate = T) %>% 
add_glance_table(
  include = c(logLik, concordance)
) %>%
  modify_caption("**Table 22: Stratified-Extended Cox Model**") 
```


$$h_g(t) = {h_{0g}(t)} \times \exp(0.0988\text{(treat-test)}-0.0464\text{(Karn*tgroup1)}+0.0097\text{(Karn*tgroup2)}-0.0179\text{(Karn*tgroup3)}$$
where $$g = 1, 2, 3, 4 \text{ (number of cell types)}$$
$$tgroup1 = 1 \text{ if  } 0 \le t < 90 $$
$$tgroup2 = 1 \text{ if  } 90 \le t < 180 $$
$$tgroup3 = 1 \text{ if  } t \ge 180$$

### 3.5. Interpretations

Based on Cox PH for after 180 days, a patient receiving the test
treatment has 0.73 times the hazard of a patient receiving the standard
treatment. In other words, a male VA patient receiving the standard
treatment is 37% more likely to die than one receiving the test
treatment when adjusting for the other predictors. However, these
results are insignificant especially when considering the small sample
size.

Based on the Cox model stratified on cell type, a patient receiving the
test treatment is 33% more likely to die than a patient receiving the
standard treatment, but these treatment results are insignificant.

Using the Stratified-Extended Cox model, a patient on the test treatment
is 10% more likely to die than one receiving the standard when adjusting
for cell type and Karnofsky score. However, the treatment results are
insignificant. The model also showed that the probability of dying for a
male VA patient with a 10-point lower Karnofsky score is 57% more likely
to die than a patient with a 10-point higher score for the first 90 days.
The Karnofsky scores across the other time intervals are
insignificant.

Overall, it seems that Karnofsky scores and cell type types are the
strongest predictors of survival. The initial Karnofsky score have
significant predictive effects at the beginning of study but the effect
diminishes as time increases. The effect of cell type also varies as
time increases. Treatment seems insignificant based on the multiple
attempted models in this analysis.

## 4. Conclusion

In this paper, we were able to review which variables were significant
in terms of impact on survival time while also ensuring that the
Proportional Hazard (PH) Assumption was upheld. In our example, we
tested various iterations of the Cox Proportional Hazard model to
determine which variables, specifically treatment, were significant in
the survival time of any given patient. The testing ranged from the
standard Cox PH model itself, to utilizing Schoenfeld's residuals and
the stratified methods, as well as removing certain variables. We
determined overall that treatment was insignificant and instead found
that through the Stratified-Extended Cox model, the Karnofsky's scores were stronger predictors of the survival time of any given patient. While it is preferable to conduct a study with more
available data, the Cox Proportional Hazards model allows some room for
censored and missing information while not compromising much in terms of
accuracy.

The Cox Proportional Hazards model is a versatile tool to analyze
covariates in the healthcare field. This, coupled with its ability to be
adjusted as necessary to a censored or limited dataset, showcases its
robust nature. With this model, researchers are able to accurately
determine which risk factors may play a part in the survival time of a
patient.

## 5. References
